From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study

### 2025-05-20 draft_v1/draft_v1.qmd
draft update

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates the dynamic relationship between precarious life conditions and depression by combining constraint-based causal discovery with computational modeling. Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between financial stress, domain-specific precariousness indicators, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precarity under varying levels of financial stress reduction. Simulation results revealed that the system's responsiveness depends not only on the intensity of external intervention but also on internal structural features such as feedback strength and noise. In particular, some configurations exhibited bistability, requiring interventions to surpass a critical threshold to trigger lasting improvement, while others responded more proportionally.
  Together, these findings emphasize that internal feedback dynamics—not just external adversity—can shape the impact of mental health interventions. More broadly, our study offers a scalable framework for integrating causal discovery and dynamical simulation to better understand and inform policy responses to complex psychosocial phenomena.

@@

This study aims to investigate the causal mechanisms linking precariousness and depression—and how those mechanisms shape system responses to external intervention. Specifically, we ask: *How does the internal configuration of the precarity–depression relationship affect the outcomes of stress-reducing interventions?*
To answer this, we combine two approaches. First, we use cycle-capable causal discovery algorithms to infer directional relationships among financial stress, domain-specific precariousness factors, and depressive symptoms. By analyzing both aggregate scores (e.g., PHQ-9 total, composite precarity index) and disaggregated variables (individual symptoms and life-domain indicators), we identify both broad associations and fine-grained causal pathways.
Second, we translate these insights into a computational dynamical model that simulates how depression and precarity co-evolve under varying levels of external stress. This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. In some configurations, the system exhibits bistability, remaining stuck in a high-risk state unless external support exceeds a critical threshold. In others, it shows unistable behavior, responding more gradually to incremental changes.
By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precarity system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study, which has been described in detail elsewhere [@stronks2013unravelling; @snijder2017cohort]. It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and which provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement]. To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
To ensure a robust representation of precariousness in our causal discovery models, we conducted various exploratory analyses to identify consistent and meaningful data-driven factor structures. Based on these analyses, we identified five precariousness factors, including two related to recent stressors, each comprising multiple variables as outlined below. DFurther details on the HELIUS study and factor extraction are provided in the [Appendix](#sec-appendix).

-   Employment precariousness: `emp_stat`, `work_sit`.
-   Social precariousness: `soc_freq`, `soc_adq`.
-   Housing precariousness: `nb_safe`, `nb_res`, `nb_rent`, `cul_rec`.
-   Recent relational stressors: `frd_brk12`, `conf12`.
-   Recent financial stressors: `fincri12`, `inc_diff`.

After preprocessing, the HELIUS dataset comprises 21,628 samples. Along with the five precariousness factors, we also compute an overall precarity score as the combined value of these factors. Additionally, PHQ-9 scores are included to represent depression, both as a total sum score and as individual symptom scores. In the subsequent causal discovery analysis, we examine the relationship between depression and precarity using both aggregated sum scores and their individual-level representations. Refer to @fig-dist for the overall distributions of the variables used in the analysis.

@@

To uncover directional relationships between precariousness and depression, we used constraint-based causal discovery algorithms that allow for feedback loops and latent confounding. Specifically, we applied the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019]. For reference, we also included the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].

A key challenge in applying these algorithms to the HELIUS dataset is that many variables are non-Gaussian and their relationships are likely nonlinear. To address this, we supplemented the commonly used Gaussian conditional independence (CI) test—based on partial correlations—with a non-parametric alternative: the *Randomized Conditional Correlation Test* (RCoT). RCoT approximates kernel-based CI testing using random Fourier features, significantly reducing computational demands while maintaining sensitivity to nonlinear dependencies [@strobl2019approximate; @zhang2012kernel].

We examined the causal structure using three complementary approaches:

@@

To ensure robust estimation and guard against sensitivity to parameter settings, we implemented a systematic grid of analysis conditions:

* Significance levels: $\alpha$ = 0.01 and 0.05
* Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8
* CI tests: Gaussian and RCoT
* Algorithms: FCI, CCI, and PC

Each combination was bootstrapped 100 times for sum-score level analyses and 30 times for symptom-level analyses, yielding a total of 1,600 graphs per algorithm for the full analysis. For symptom networks, we fixed the skeleton structure using a consensus graph estimated across PC, FCI, and CCI ($\alpha$ = 0.01, RCoT). This constraint improves computational efficiency without limiting the estimation of directional edges between symptoms and precariousness variables.

To summarize the results, we adopted a two-level aggregation strategy. First, within each analysis condition, we recorded the most frequently observed edge-endpoint type (e.g., directed, undirected, uncertain) from the bootstraps. Then, across all conditions, we tallied the most frequent endpoint type per edge. In cases of ties, we assigned dashed edges to reflect uncertainty. For transparency, we accompany each graph with a table showing the relative frequencies of each edge symbol.

This procedure yields a final graph for each algorithm and analysis type—offering both robustness and interpretability. See @fig-workflow for an overview of the causal discovery pipeline.

For detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, see the [Appendix](#sec-appendix).

@@

```{r}
#| label: fig-workflow
#| fig-cap: "Analysis workflow applied across all three algorithms."
#| fig-align: center
#| out-width: 80%

knitr::include_graphics("img/simsetup2.png")

@@

```{r}
#| label: fig-concept
#| fig-cap: "Conceptual high-level causal structure of precariousness and depression system."
#| fig-align: center
#| out-width: 35%

@@

To identify realistic parameter settings for the simulation model, we calibrated six free parameters—$\alpha_1$, $\alpha_2$, $\beta_1$, $\beta_2$, and noise amplitudes $\sigma_1$, $\sigma_2$—using the Non-dominated Sorting Genetic Algorithm II (NSGA-II) [@deb2002fast], a multi-objective evolutionary optimization method well-suited for noisy and high-dimensional search spaces. The goal was to match simulated system behavior to empirical summary statistics derived from the HELIUS dataset, including the means and variances of depression ($D$) and social precariousness ($P$), as well as their partial correlations after adjusting for financial stress ($S$), specifically $\rho_{DP.S}$, $\rho_{DS.P}$, and $\rho_{PS.D}$. 
Given the stochastic nature of the model, each candidate parameter set was evaluated by simulating 1,000 independent trajectories, each representing an individual with a fixed $S$ value sampled from the observed distribution. Final-state values were extracted and aggregated to compute population-level summary statistics. Fitness was assessed as the absolute deviation from the empirical targets, aggregated across all objectives. 
NSGA-II was implemented via the nsga2R package [@nsga2R], with a population of 300 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2). This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives. From the resulting Pareto front (@fig-pareto), we selected the top 30 parameter sets with the lowest combined error to use in the subsequent simulations assessing intervention effects.

@@

For each of the top 30 calibrated parameter sets, we simulate the system’s evolution across 300 individuals under a continuum of intervention levels. Each simulation run generates final values for $D$ and $P$, which are then used to assess how internal model parameters influence the system’s behavior under changing external conditions.

@@

The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (PHQsum). Employment precarity (P.emp) and social precarity (P.soc) are not identified as causes of depression, whereas financial stress (S.fin) appears to play a causal role. While P.emp and P.soc are generally not recognized as causes of other precariousness factors, S.fin emerges as a potential cause, as indicated by its circle edge endpoint. Supporting this, the proportion matrix plot in panel (b) shows that S.fin has some probability of causally influencing either P.emp or P.soc.

@@

Relational stress (S.rel) plays a more nuanced role, interacting with depression and S.fin  through a latent confounder. Its relationship with P.soc remains less certain, as the proportion table suggests a fair probability that this connection is absent, uncertain, or bidirectional. Meanwhile, housing precarity (P.hou) is not directly causally related to depression but is linked to P.emp and S.fin. While P.emp and S.fin are identified as non-causes of P.hou, it remains unclear whether P.hou causally influences P.emp or if their relationship is mediated by an unobserved confounder.

The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify S.fin as a cause of depression. Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI. Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.

@@

Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms. Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph highlights heterogeneity in how precariousness factors influence individual depressive symptoms: con (concentration), slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal), anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.

The symptom-level graph in @fig-sym panel (a) reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges. As before, dashed edges denote ties, with specific ties marked by bold horizontal bars. 
In @fig-sym panel (b), the proportion matrix provides more insight into these directionalities. Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type. However, blue regions, which correspond to circle endpoints, are particularly common for anh, slp, ene, and sui, reflecting uncertainty in the causal direction among these symptoms. Notably, the only arrowtail connection appears between dep and anh, suggesting that anhedonia (anh) is a potential cause of depressed mood (dep).

Looking at the symptom-precariousness connections, one of the key patterns is that ties are most frequently found in the relationships between individual symptoms and precariousness factors. This suggests that these causal links may be less stable across different conditions. Upon closer examination, most ties arise due to discrepancies between the Gaussian CI test and the RCoT test—with Gaussian CI favoring arrowheads and RCoT more often predicting the absence of an edge.

This discrepancy likely arises from a combination of factors. 
First, as the analysis shifts from aggregated to disaggregated variables, the number of conditional independence (CI) tests increases, leading to a loss of statistical power in high-dimensional causal discovery. 
With more variables being conditioned on, the data becomes sparser, making it harder to detect weaker dependencies. Additionally, over-conditioning—controlling for too many variables—can artificially remove statistical associations, leading to false negatives.
Beyond this general statistical power issue, the fundamental differences in how these CI tests operate contribute to the inconsistency. RCoT, as a nonparametric test, does not assume linearity and can capture both linear and nonlinear dependencies. However, it is more conservative and requires larger sample sizes, making it more prone to false negatives in high-dimensional settings. 
In contrast, partial correlation (Gaussian CI test) assumes linear Gaussian relationships, making it more permissive and sometimes detecting weak dependencies that may not be statistically meaningful. 
As a result, when Gaussian CI detects an edge while RCoT does not, it could indicate that the relationship is weak, strictly linear, or requires more data for reliable nonparametric detection.
This explains why many ties in the proportion matrix show a 50-50 split between symbols—where half of the conditions with RCoT suggest the absence of an edge, while the other half using Gaussian CI test favor an arrowhead. This systematic divergence highlights the challenges of causal discovery in high-dimensional settings, where varying statistical assumptions can lead to differing conclusions about the presence and direction of causal relationships.

Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of financial stress (S.fin), which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality. One exception is the stronger evidence suggesting that S.fin may cause changes in app (appetite), while its relationships with other symptoms, such as ene, mot, and dep, remain ambiguous.
Similarly, social precarity (P.soc) exhibits numerous connections with depressive symptoms, with most edges pointing toward P.soc rather than outward from it. This suggests that depressive symptoms, particularly dep and glt, may contribute to worsening social precarity rather than the other way around. Additionally, anh also shows some probability of causally influencing P.soc, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.

Other precariousness factors exhibit more uncertain but still notable relationships. Relational stress (S.rel) shows weak but existing connections with slp, glt, and app, though directionality remains unclear in many cases. Employment precarity (P.emp) also connects to symptoms like slp, mot, and con, but, like S.rel, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
In line with the aggregated graph in @fig-sum, housing precarity (P.hou) does not appear to have any direct relationship with depression symptoms but consistently shows associations with employment precarity (P.emp) and financial stress (S.fin). The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with financial stress (S.fin) exhibiting the strongest tendency to influence other precariousness factors.

>>>>

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates the dynamic relationship between precarious life conditions and depression by combining constraint-based causal discovery with computational modeling. Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between domain-specific precariousness indicators, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precarity under varying levels of financial stress reduction. Simulation results revealed that the system's responsiveness depends not only on the intensity of external intervention but also on internal structural features such as feedback strength and noise. In particular, some configurations exhibited bistability, requiring interventions to surpass a critical threshold to trigger lasting improvement, while others responded more proportionally.
  Together, these findings emphasize that internal feedback dynamics—not just external adversity—can shape the impact of mental health interventions. More broadly, our study offers a scalable framework for integrating causal discovery and dynamical simulation to better understand and inform policy responses to complex psychosocial phenomena.

@@

This study aims to investigate the causal mechanisms linking precariousness and depression—and how those mechanisms shape system responses to external intervention. Specifically, we ask: *How does the internal configuration of the precarity–depression relationship affect the outcomes of stress-reducing interventions?*
To answer this, we combine two approaches. First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors, and depressive symptoms.
By examining both aggregated measures (depression sum score and a composite precarity index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
Second, we translate these insights into a computational dynamical model that simulates how depression and precarity co-evolve under varying levels of external stress. This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. In some configurations, the system exhibits bistability, remaining stuck in a high-risk state unless external support exceeds a critical threshold. In others, it shows unistable behavior, responding more gradually to incremental changes.
By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precarity system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study, which has been described in detail elsewhere [@stronks2013unravelling; @snijder2017cohort]. It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and which provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement]. To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures. These analyses led to the identification of five distinct precariousness factors—three domain-based and two reflecting recent stressors—each composed of multiple observed variables, as outlined below. Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).

-   Employment precariousness: `emp_stat` (current employment status), `work_sit` (nature of work situation).
-   Social precariousness: `soc_freq` (frequency of social contact), `soc_adq` (perceived adequacy of social support).
-   Housing precariousness: `nb_safe` (neighborhood safety), `nb_res` (access to neighborhood resources), `nb_rent` ( proportion of rental housing), `cul_rec` (availability of cultural resources).
-   Recent relational stressors: `frd_brk12` (experience of a friendship breakup), `conf12` (recent interpersonal or relational conflict).
-   Recent financial stressors: `fincri12` (experience of a financial crisis), `inc_diff` ( difficulty managing household income).

After preprocessing, the HELIUS dataset includes 21,628 samples. In addition to the five identified precariousness factors, we computed an overall precarity score by aggregating their values. Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores. In the subsequent causal discovery analysis, we explore the relationship between depression and precarity using both aggregated scores and their disaggregated components. See @fig-dist for the overall distributions of the variables used in the analysis.

@@

To uncover directional relationships between precariousness and depression, we use constraint-based causal discovery algorithms that allow for feedback loops and latent confounding. Specifically, we apply the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019]. 
For reference, we also include the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].
While our full analysis includes results from three causal discovery algorithms—FCI, CCI, and PC—we focus on FCI in the main text, as it offers a strong balance between flexibility and theoretical soundness. FCI accommodates latent confounding and potential feedback, making it well-suited for the complexity of our data [@mooij2020constraint]. The PC algorithm is included for reference but relies on more restrictive assumptions, such as acyclicity and no unmeasured confounding. CCI, while capable of modeling cycles explicitly, comes with certain theoretical limitations. Results from both PC and CCI are provided in the [Appendix](#sec-appendix) ([@sec-cci; @sec-pc]) for completeness and comparison. For readers interested in detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, please refer to the [Appendix](#sec-appendix) ([@sec-causalprimer]).

A key challenge in applying these algorithms to the HELIUS dataset is that many variables are non-Gaussian and likely exhibit nonlinear relationships. While kernel-based conditional independence tests (such as KCIT) are well-suited for capturing such complex dependencies, they are computationally intensive—scaling quadratically with sample size due to the need to invert large kernel matrices—making them impractical for large datasets like HELIUS [@rahimi2007random].
To address this, we use the Randomized Conditional Correlation Test (RCoT), a nonparametric alternative that approximates kernel methods using random Fourier features. This reduces computational complexity from quadratic to linear in sample size, substantially lowering runtime while preserving sensitivity to nonlinear relationships [@strobl2019approximate; @zhang2012kernel]. For further details, see @sec-rcot.

We examine the causal structure using three complementary approaches:

@@

To robustly estimate the causal structure between depression and precariousness variables, we implement a comprehensive bootstrapped causal discovery procedure. This procedure systematically varies key analysis parameters across a grid of configurations, enabling us to assess the stability of inferred relationships under different settings.
Specifically, we vary four main components:

* Significance levels for conditional independence tests: $\alpha$ = 0.01 and 0.05.
* Stability thresholds: 0.5, 0.6, 0.7, and 0.8, which determine how frequently an edge must appear across bootstraps to be retained.
* Conditional independence (CI) tests: Gaussian CI test (partial correlation) and RCoT (a nonparametric test).
* Causal discovery algorithms: FCI, CCI, and PC.

For each unique parameter combination (2 significance levels × 4 thresholds × 2 CI tests = 16 settings), we run 100 bootstrap samples per algorithm for the aggregated analyses (see Step 1 in @fig-workflow). For symptom-level analyses, which are more computationally intensive, we use 30 bootstrap samples per setting.[^5]

[^5]: For symptom-level analyses, we further improve computational efficiency by fixing the skeleton structure—the undirected network of potential connections—using a consensus graph derived from all three algorithms (with $\alpha = 0.01$ and the RCoT test). This allows subsequent analyses to focus exclusively on estimating edge directions, avoiding the repeated and costly computation of skeletons, and thus substantially reducing runtime.

To summarize these results, we apply a two-level aggregation strategy:

1. **Within each condition** (e.g., fixed $\alpha$, threshold, CI test, and algorithm), we identify the most frequently occurring edge type (e.g., arrowhead, circle, none) across bootstraps for each pair of variables (Step 2 in @fig-workflow).
2. **Across conditions**, we then determine the most dominant edge type based on majority voting. In cases where no edge type is clearly dominant (i.e., a tie), we mark the edge as uncertain, represented with a dashed line in the final graph (Step 3 in @fig-workflow).

This summarization process is repeated separately for each causal discovery algorithm (FCI, CCI, PC), yielding a single robust graph per algorithm. These final graphs represent only the most stable and consistent relationships across a wide range of plausible parameterizations, enhancing both reliability and interpretability.
To support transparency, each final graph is accompanied by a matrix summarizing the relative frequency of each edge type for every variable pair—enabling readers to assess the degree of uncertainty or agreement underlying each connection.

<!-- To ensure robust estimation and guard against sensitivity to parameter settings, we implemented a systematic grid of analysis conditions: -->

<!-- * Significance levels: $\alpha$ = 0.01 and 0.05 -->
<!-- * Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8 -->
<!-- * CI tests: Gaussian and RCoT -->
<!-- * Algorithms: FCI, CCI, and PC -->

<!-- Each combination was bootstrapped 100 times for sum-score level analyses and 30 times for symptom-level analyses, yielding a total of 1,600 graphs per algorithm for the full analysis. For symptom networks, we fixed the skeleton structure using a consensus graph estimated across PC, FCI, and CCI ($\alpha$ = 0.01, RCoT). This constraint improves computational efficiency without limiting the estimation of directional edges between symptoms and precariousness variables. -->

<!-- To summarize the results, we adopted a two-level aggregation strategy. First, within each analysis condition, we recorded the most frequently observed edge-endpoint type (e.g., directed, undirected, uncertain) from the bootstraps. Then, across all conditions, we tallied the most frequent endpoint type per edge. In cases of ties, we assigned dashed edges to reflect uncertainty. For transparency, we accompany each graph with a table showing the relative frequencies of each edge symbol. -->

<!-- This procedure yields a final graph for each algorithm and analysis type—offering both robustness and interpretability. See @fig-workflow for an overview of the causal discovery pipeline. -->
<!-- For detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, see the [Appendix](#sec-appendix). -->

@@

```{r}
#| label: fig-workflow
#| fig-cap: "Causal discovery workflow applied across all three algorithms (FCI, CCI, and PC). *The process involves three main steps: (1) bootstrapping the dataset to create multiple resampled datasets; (2) estimating causal graphs under varying parameter configurations; and (3) summarizing results by identifying the most frequent edge type across bootstraps and settings to construct a final stable graph.*"
#| fig-align: center
#| out-width: 90%

knitr::include_graphics("img/simsetup.pdf")

@@

```{r}
#| label: fig-concept
#| fig-cap: "High-level conceptual model of causal structure linking social precariousness (P.soc) and depression (PHQsum), influenced by external financial stress (S.fin)."
#| fig-align: center
#| out-width: 40%

@@

To identify realistic parameter settings for the simulation model, we calibrate six free parameters—$\alpha_1$, $\alpha_2$, $\beta_1$, $\beta_2$, and noise amplitudes $\sigma_1$, $\sigma_2$—using the *Non-dominated Sorting Genetic Algorithm II* (NSGA-II) [@deb2002fast], a multi-objective evolutionary optimization method well-suited for noisy and high-dimensional search spaces. 
The goal is to match simulated system behavior to empirical summary statistics derived from the HELIUS dataset. Specifically, we aim to match the observed means and variances of depression ($D$) and social precariousness ($P$), along with all three partial correlations among $D$, $P$, and financial stress ($S$): $\rho_{DP.S}$, $\rho_{DS.P}$, and $\rho_{PS.D}$.

Given the stochastic nature of the model, each candidate parameter set is evaluated by simulating 1,000 virtual individuals, each assigned a fixed $S$ value sampled from the empirical distribution. From these simulations, we compute population-level summary statistics based on the final states. The fitness of each parameter set is measured by how closely its summary statistics matched the empirical data, using the total absolute deviation across all objectives.

NSGA-II is implemented via the `nsga2R` package [@nsga2R], with a population of 500 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2). This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives. From the resulting Pareto front (@fig-pareto), we selected the top 30 parameter sets with the lowest combined error to use in the subsequent simulations assessing intervention effects.

@@

For each of the top 30 parameter sets identified through calibration, we run simulations on 300 virtual individuals across this full range of intervention levels. In each simulation, we observe how depression ($D$) and precariousness ($P$) evolve over time, capturing their final values after the system stabilizes. This setup allows us to evaluate how different internal model dynamics (e.g., feedback strength) influence the system’s response to external change.

@@

The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (*PHQsum*). Employment precarity (*P.emp*) and social precarity (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role. While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint. Supporting this, the proportion matrix plot in panel (b) shows that *S.fin* has some probability of causally influencing either *P.emp* or *P.soc.*

@@

Relational stress (*S.rel*) plays a more nuanced role, interacting with depression and *S.fin*  through a latent confounder. Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 indicates a considerable probability that this relationship is either absent, ambiguous, or potentially bidirectional. Meanwhile, housing precarity (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*. While *P.emp* and *S.fin* are identified as non-causes of *P.hou*, it remains unclear whether *P.hou* causally influences *P.emp* or if their relationship is mediated by an unobserved confounder.

The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify *S.fin* as a cause of depression. Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI. Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.

@@

Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms. Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.

The symptom-level graph in @fig-sym-1 reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges. As before, dashed edges denote ties, with specific ties marked by bold horizontal bars. 
In @fig-sym-2, the proportion matrix provides more insight into these directionalities. Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type. However, blue regions, which correspond to circle endpoints, are particularly common for *anh*, *slp*, *ene*, and *sui*, reflecting uncertainty in the causal direction among these symptoms. Notably, the only arrowtail connection appears between *dep* and *anh*, suggesting that anhedonia is a potential cause of depressed mood.

Looking at the symptom-precariousness connections, one of the key patterns is that ties are most frequently found in the relationships between individual symptoms and precariousness factors. This suggests that these causal links may be less stable across different conditions. Closer examination reveals that most ties arise due to discrepancies between the Gaussian CI test and the RCoT test—with Gaussian CI favoring arrowheads and RCoT more often predicting the absence of an edge.

This discrepancy likely arises from a combination of factors. 
As the analysis moves from aggregated to disaggregated variables, the number of CI tests increases substantially, leading to a loss of statistical power. With more variables being conditioned on, the effective sample size per test diminishes, making it harder to detect subtle dependencies. Over-conditioning—controlling for too many variables—can also obscure real relationships, resulting in false negatives.
Beyond these statistical power issues, the underlying assumptions of the two CI tests contribute to their divergence. RCoT is a nonparametric test that captures both linear and nonlinear dependencies without assuming a specific distributional form. However, it is more conservative and requires larger sample sizes, especially in high-dimensional settings. In contrast, the Gaussian CI test is based on partial correlations and assumes linear Gaussian relationships, making it more permissive and more likely to detect weak or spurious associations.
As a result, it is common to see edges where Gaussian CI suggests a directional relationship, while RCoT indicates no edge. In the proportion matrix, this appears as a 50–50 tie between absence and arrowhead symbols. This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related.

Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of *S.fin*, which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality. One exception is the stronger evidence suggesting that *S.fin* may cause changes in *app*, while its relationships with other symptoms, such as *ene*, *mot*, and *dep*, remain ambiguous.
Similarly, *P.soc* exhibits numerous connections with depressive symptoms, with most edges pointing toward *P.soc* rather than outward from it. This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precarity rather than the other way around. Additionally, *anh* also shows some probability of causally influencing *P.soc*, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.

Other precariousness factors exhibit more uncertain but still notable relationships. *S.rel* shows weak but existing connections with *slp*, *glt*, and *app*, though directionality remains unclear in many cases. *P.emp* also connects to symptoms like *slp*, *mot*, and *con*, but, like *S.rel*, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
In line with the aggregated graph in @fig-sum, *P.hou* does not appear to have any direct relationship with depression symptoms but consistently shows associations with *P.emp* and *S.fin*. The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with *S.fin* exhibiting the strongest tendency to influence other precariousness factors.

Among depressive symptoms, *dep*, *glt*, and *slp* appear to be the most connected to precariousness factors, while *P.soc* has the most connections with symptoms, predominantly as a recipient rather than a driver of influence. Within the symptom network, *slp*, *sui*, and *anh* emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms. On the other hand, *con*, *mot*, *dep*, and *glt*, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
Considering both symptom-precariousness connections and symptom-level dynamics, *slp* emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network. This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precarity. On the other hand, *P.soc* serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.
